Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems

### 2024-06-15 final/oup-authoring-template.tex
Spellchecing

One limitation of these tools is their extensibility from Python.
As they wrap libraries in other languages for defining, translating and/or solving models,
these steps cannot be altered or easily inspected from Python.
While they enable model definition and running simulations via Python scripts,
they cannot fully leverage Python's extensive package ecosystem.
For example,
COPASI and Tellurium do not allow the use of solvers defined in other Python packages,
and adding new integrators requires working with C++.
In particular, the step that translates into equations is not exposed by any of these tools.
As such, it is not possible to apply custom optimizations to the equations
or use automatic differentation packages such as JAX \citep{jax2018github} to compute the model's jacobian.

@@

To overcome these limitations,
we developed poincaré and SimBio,
open-source Python packages for defining, translating and solving systems.
Poincaré allows one to define differential equation systems
using variables, parameters and constants,
and assiging rate equations to variables.
For defining \acp{CRN},
SimBio builds on top of poincaré providing species and reactions that keep track of stoichiometries.
Both are focused on providing an ergonomic experience to end-users
by integrating well with \acp{IDE} and static analysis tools
through the use of standard modern Python syntax.
Moreover, since they are coded in pure Python,
each step from model definition, translation to equations or solving
can be extended or debugged from Python.
Being the first-ever pure Python packages for systems modelling,
they offer extensive extensibility,
from simple tasks like reusing integrators defined in other packages,
to complex ones like altering the compilation process to leverage some structure in the equations.
For example, using a for-loop in the compiled equations could improve the runtime performance
if there is some repetitive structure in the system,
as happens in spatial modelling.
The models built using these packages can be introspected to create other representations,
such as graphs connecting species and/or reactions, or tables with parameters or equations.
Furthermore, they have a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.
We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
We also highlighted the substantial performance improvements our tools offer,
as this is crucial for construction and simulation of models of whole cells and organisms,
which necessitate the simulation of significantly large-scale models.

@@

Poincare is a package to define and simulate dynamical systems.
Based on Python inmutable dataclasses,
it provides a \texttt{System} class,
where one can define
\texttt{Constant}s; which can be numbers or refer to other constants;
\texttt{Parameter}s, which can be numbers or time-dependent expressions;
\texttt{Variable}s, which represent the state of the system and must be provided with an initial condition;
and create equations linking a variable's derivative with an expression (Figure~\ref{fig-first-order}).
It also allows to
create an \texttt{Independent} variable to define non-autonomous systems,
and define higher-order systems by assigning an initial condition to a \texttt{Derivative} (Figure~\ref{fig-second-order}).

@@

% FIGURE 6
\begin{figure*}[t]
  \centering
  \includegraphics[width=\textwidth]{performance}
  \caption{
    Performance of different softwares to solve models from the curated section of BioModels.
    (left) Run time for the model BIOMD3 as a function of the number of output points.
    (right) Run time for different models for 300 output points,
    using the geometric mean of the different softwares to order them.
    Each point corresponds to the median of 20 runs,
    with a neglibible errorbar given by the interquartile range.
  }
  \label{fig-runtime}
\end{figure*}

@@

For COPASI and Tellurium,
we noted that its runtime depended on the number of intermediate evaluation points returned to the user (Figure~\ref{fig-runtime}, left).
We speculate that this is due to memory allocation and data transfer in the Python bindings, 
as the number of total function evaluations (i.e. including those performed by the integrator steper)
is around 3000 and therefore much larger than most part of the x-axis.
For SimBio,
its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
Nevertheless, switching to the numba backend,
which \ac{JIT} compiles the \ac{RHS} equations,
puts it on par with them.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
between each of the integration steps.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
For the models considered,
amortizing the compilation time required from 2 upto 200 runs.

>>>>

One limitation of these tools is their extensibility from Python.
As they wrap libraries in other languages for defining, translating and/or solving models,
these steps cannot be altered or easily inspected from Python.
While they enable model definition and running simulations via Python scripts,
they cannot fully leverage Python's extensive package ecosystem.
For example,
COPASI and Tellurium do not allow the use of solvers defined in other Python packages,
and adding new integrators requires working with C++.
In particular, the step that translates into equations is not exposed by any of these tools.
As such, it is not possible to apply custom optimizations to the equations
or use automatic differentiation packages such as JAX \citep{jax2018github} to compute the model's jacobian.

@@

To overcome these limitations,
we developed poincaré and SimBio,
open-source Python packages for defining, translating and solving systems.
Poincaré allows one to define differential equation systems
using variables, parameters and constants,
and assigning rate equations to variables.
For defining \acp{CRN},
SimBio builds on top of poincaré providing species and reactions that keep track of stoichiometries.
Both are focused on providing an ergonomic experience to end-users
by integrating well with \acp{IDE} and static analysis tools
through the use of standard modern Python syntax.
Moreover, since they are coded in pure Python,
each step from model definition, translation to equations or solving
can be extended or debugged from Python.
Being the first-ever pure Python packages for systems modelling,
they offer extensive extensibility,
from simple tasks like reusing integrators defined in other packages,
to complex ones like altering the compilation process to leverage some structure in the equations.
For example, using a for-loop in the compiled equations could improve the runtime performance
if there is some repetitive structure in the system,
as happens in spatial modelling.
The models built using these packages can be introspected to create other representations,
such as graphs connecting species and/or reactions, or tables with parameters or equations.
Furthermore, they have a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.
We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
We also highlighted the substantial performance improvements our tools offer,
as this is crucial for construction and simulation of models of whole cells and organisms,
which necessitate the simulation of significantly large-scale models.

@@

Poincare is a package to define and simulate dynamical systems.
Based on Python immutable dataclasses,
it provides a \texttt{System} class,
where one can define
\texttt{Constant}s; which can be numbers or refer to other constants;
\texttt{Parameter}s, which can be numbers or time-dependent expressions;
\texttt{Variable}s, which represent the state of the system and must be provided with an initial condition;
and create equations linking a variable's derivative with an expression (Figure~\ref{fig-first-order}).
It also allows to
create an \texttt{Independent} variable to define non-autonomous systems,
and define higher-order systems by assigning an initial condition to a \texttt{Derivative} (Figure~\ref{fig-second-order}).

@@

% FIGURE 6
\begin{figure*}[t]
  \centering
  \includegraphics[width=\textwidth]{performance}
  \caption{
    Performance of different softwares to solve models from the curated section of BioModels.
    (left) Run time for the model BIOMD3 as a function of the number of output points.
    (right) Run time for different models for 300 output points,
    using the geometric mean of the different softwares to order them.
    Each point corresponds to the median of 20 runs,
    with a negligible error-bar given by the interquartile range.
  }
  \label{fig-runtime}
\end{figure*}

@@

For COPASI and Tellurium,
we noted that its runtime depended on the number of intermediate evaluation points returned to the user (Figure~\ref{fig-runtime}, left).
We speculate that this is due to memory allocation and data transfer in the Python bindings, 
as the number of total function evaluations (i.e. including those performed by the integrator stepper)
is around 3000 and therefore much larger than most part of the x-axis.
For SimBio,
its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
Nevertheless, switching to the numba backend,
which \ac{JIT} compiles the \ac{RHS} equations,
puts it on par with them.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
between each of the integration steps.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
For the models considered,
amortizing the compilation time required from 2 upto 200 runs.
